New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions

### 2022-07-11 paper/paper.Rmd
projections 4.2

To illustrate the usefulness of the manual tour we use the 4D penguins data [@penguins]. We will show how classification boundaries can be explored and better understood using a manual tour, on projections and slices through 4D space. Figure \ref{fig:penguins-scatmat} shows a scatterplot matrix of this data. There are four variables (`bl = bill_length_mm, bd = bill_depth_mm, fl = flipper_length_mm, bm = body_mass_g`) measuring the size of the penguins from three species (Adelie, Chinstrap and Gentoo). The scatterplot matrix shows that the three species appear to be likely separable, and that at least the Gentoo can be distinguished from the other two species when `bd` is paired with `fl` or `bm`. The steps for exploring boundaries are as follows:

1. Build classification model, here we use a random forest and linear discriminant analysis.
2. Predict the class for a dense grid of values covering the data space.
3. Examine projections, using a manual tour so that the contribution of any variable is controlled.
4. Slice through the center, to explore where the boundaries will likely meet.
5. Move the slice by changing the center in the direction of a single variable to explore the extent of a boundary for a single group relative to a variable.

@@

We use the `classifly` package [@classifly] to generate predictions across the 4D cube spanned by the data, with two classification models: linear discriminant analysis (LDA) and random forest(RF).

@@

We start by first exploring the projections of the model prediction. Through manual rotation of the view we can get a feeling for where in the space we primarily predict each of the three species, and we also get a sense of the difference between the two models. To illustrate this difference we have rotated the projection view from the RF model to identify a projection that shows the non-linear, block-type structure introduced by the model, see Fig. \ref{proj1}. We then export the corresponding matrix $A_1$ to look at the same projection for the LDA model and the original data. Here we notice that while the overall picture is similar, the linearity of the LDA leads to a big difference in how two of the three species overlap across the projection (blue vs green, i.e. Adelie vs Gentoo penguins). Interestingly that same projection leads to almost perfect separation of the three species when looking at the data points, the overlap between the models is driven by what is happening in the orthogonal space to the plane.

>>>>

To illustrate the usefulness of the manual controls we use the 4D penguins data [@penguins]. We will show how classification boundaries can be explored and better understood on projections and slices through 4D space. Figure \ref{fig:penguins-scatmat} shows a scatterplot matrix of this data. There are four variables (`bl = bill_length_mm, bd = bill_depth_mm, fl = flipper_length_mm, bm = body_mass_g`) measuring the size of the penguins from three species (Adelie, Chinstrap and Gentoo). The scatterplot matrix shows that the three species appear to be likely separable, and that at least the Gentoo can be distinguished from the other two species when `bd` is paired with `fl` or `bm`. The steps for exploring boundaries in this example are as follows:

1. Build your classification model. 
2. Predict the class for a dense grid of values covering the data space.
3. Examine projections, using a manual tour so that the contribution of any variable is controlled.
4. Slice through the center, to explore where the boundaries will likely meet.
5. Move the slice by changing the center in the direction of a single variable to explore the extent of a boundary for a single group relative to a variable.

@@

We use the `classifly` package [@classifly] to generate predictions across the 4D cube spanned by the data, with two classification models: linear discriminant analysis (LDA) and random forest (RF).

@@

We start by first exploring the projections of the model prediction. Figure \ref{proj1} summarises the process. Through manual rotation of the view we can get a feeling for where in the space we primarily predict each of the three species, and we also get a sense of the difference between the two models. To illustrate this difference we have manually rotated the projection for the RF model (left plot) to identify a projection that shows the non-linear but block-type structure that is typical of this type of model. This particular projection ($A_1$) is exported so that the same projection can be used to show the LDA model (middle plot) and the actual data (right plot). What can be seen is the linearity of the LDA model, where the boundaries are linear and oblique to the variable axes. And, interestingly this particular projection of the original data shows very distinct clusters of the three species. That means, the obscuring of the boundaries between groups for both of the models is driven by what is happening in the orthogonal space to the plane of the selected projection.

### 2022-07-11 paper/paper.tex
projections 4.2

To illustrate the usefulness of the manual tour we use the 4D penguins
data \citep{penguins}. We will show how classification boundaries can be
explored and better understood using a manual tour, on projections and
slices through 4D space. Figure \ref{fig:penguins-scatmat} shows a
scatterplot matrix of this data. There are four variables
(\texttt{bl\ =\ bill\_length\_mm,\ bd\ =\ bill\_depth\_mm,\ fl\ =\ flipper\_length\_mm,\ bm\ =\ body\_mass\_g})
measuring the size of the penguins from three species (Adelie, Chinstrap
and Gentoo). The scatterplot matrix shows that the three species appear
to be likely separable, and that at least the Gentoo can be
distinguished from the other two species when \texttt{bd} is paired with
\texttt{fl} or \texttt{bm}. The steps for exploring boundaries are as
follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Build classification model, here we use a random forest and linear
  discriminant analysis.
\item
  Predict the class for a dense grid of values covering the data space.
\item
  Examine projections, using a manual tour so that the contribution of
  any variable is controlled.
\item
  Slice through the center, to explore where the boundaries will likely
  meet.
\item
  Move the slice by changing the center in the direction of a single
  variable to explore the extent of a boundary for a single group
  relative to a variable.
\end{enumerate}

@@

@@

We use the \texttt{classifly} package \citep{classifly} to generate
predictions across the 4D cube spanned by the data, with two
classification models: linear discriminant analysis (LDA) and random
forest(RF).

@@

We start by first exploring the projections of the model prediction.
Through manual rotation of the view we can get a feeling for where in
the space we primarily predict each of the three species, and we also
get a sense of the difference between the two models. To illustrate this
difference we have rotated the projection view from the RF model to
identify a projection that shows the non-linear, block-type structure
introduced by the model, see Fig. \ref{proj1}. We then export the
corresponding matrix \(A_1\) to look at the same projection for the LDA
model and the original data. Here we notice that while the overall
picture is similar, the linearity of the LDA leads to a big difference
in how two of the three species overlap across the projection (blue vs
green, i.e.~Adelie vs Gentoo penguins). Interestingly that same
projection leads to almost perfect separation of the three species when
looking at the data points, the overlap between the models is driven by
what is happening in the orthogonal space to the plane.

>>>>

@@

To illustrate the usefulness of the manual controls we use the 4D
penguins data \citep{penguins}. We will show how classification
boundaries can be explored and better understood on projections and
slices through 4D space. Figure \ref{fig:penguins-scatmat} shows a
scatterplot matrix of this data. There are four variables
(\texttt{bl\ =\ bill\_length\_mm,\ bd\ =\ bill\_depth\_mm,\ fl\ =\ flipper\_length\_mm,\ bm\ =\ body\_mass\_g})
measuring the size of the penguins from three species (Adelie, Chinstrap
and Gentoo). The scatterplot matrix shows that the three species appear
to be likely separable, and that at least the Gentoo can be
distinguished from the other two species when \texttt{bd} is paired with
\texttt{fl} or \texttt{bm}. The steps for exploring boundaries in this
example are as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Build your classification model.
\item
  Predict the class for a dense grid of values covering the data space.
\item
  Examine projections, using a manual tour so that the contribution of
  any variable is controlled.
\item
  Slice through the center, to explore where the boundaries will likely
  meet.
\item
  Move the slice by changing the center in the direction of a single
  variable to explore the extent of a boundary for a single group
  relative to a variable.
\end{enumerate}

@@

We use the \texttt{classifly} package \citep{classifly} to generate
predictions across the 4D cube spanned by the data, with two
classification models: linear discriminant analysis (LDA) and random
forest (RF).

@@

We start by first exploring the projections of the model prediction.
Figure \ref{proj1} summarises the process. Through manual rotation of
the view we can get a feeling for where in the space we primarily
predict each of the three species, and we also get a sense of the
difference between the two models. To illustrate this difference we have
manually rotated the projection for the RF model (left plot) to identify
a projection that shows the non-linear but block-type structure that is
typical of this type of model. This particular projection (\(A_1\)) is
exported so that the same projection can be used to show the LDA model
(middle plot) and the actual data (right plot). What can be seen is the
linearity of the LDA model, where the boundaries are linear and oblique
to the variable axes. And, interestingly this particular projection of
the original data shows very distinct clusters of the three species.
That means, the obscuring of the boundaries between groups for both of
the models is driven by what is happening in the orthogonal space to the
plane of the selected projection.
